\documentclass[a4paper, 12pt]{article}

\usepackage[margin = 1in, top = 1.0in]{geometry}
\usepackage{titlesec}
\usepackage{times}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{hyperref}
%\usepackage{changepage}
\usepackage{titlesec}
\usepackage[yyyymmdd]{datetime}\renewcommand{\dateseparator}{.}
\usepackage{setspace}
\usepackage{ragged2e}

\definecolor{fade}{gray}{0.4}




\begin{document}
\pagestyle{empty}


\spacing{1}
\noindent March 9, 2021
% \vspace{0.2em}
\begin{tabbing}
\\ Shetland Space Centre
\\ Saxa Vord Resort Haroldswick, Unst, Shetland, Scotland, ZE2 9TJ
\\ Moray Park, Seafield Ave, Grantown-on-Spey PH26 3JF
\end{tabbing}



\spacing{1} \color{black}
\vspace{1em}
\noindent To whom it may concern:
\vspace{2em}

Thank you so much for considering me for a position at Shetland Space. The more I learn about your company, the more I can't help but hope that I could be part of its growth. I am currently in my final year of my BSc Physics degree at the University of St. Andrews. I believe that with my combined knowledge of physics, computer systems, and programming, I can make myself a valuable addition to your team.
At the risk of perhaps coming on too strong, for as long as I can remember, I have hoped that my future would be within the space industry. When it became up to me, through secondary school and university, I made every effort to take as many physics, math, and astrophysics classes as I could. That is why I was thrilled to be a part of Slooh.com, a website that provides users with remote access to telescopes in the Canary Islands and Chile. While I was there, the website also extended its educational capacity, which included guided tours of the basic, observable aspects of the night sky. I was able to put myself in the shoes of a student as I attempted to explain essential astronomy as elegantly as possible. I was proud to be able to teach myself the complex technologies used at Slooh within a couple of days.

\vspace{1em}
The idea of participating in the development of a satellite launch site is exhilarating and I would love the chance to apply myself and contribute to the advancement of space operations in the U.K. I will accept any job or task that is offered if it means I can be part of Shetland Space. I can offer an easygoing personality, a commitment to figuring anything out, zero need for creature comfort, a strong interest in adventure, and perhaps most importantly, a genuine desire to live and work in a remote and beautiful place.

\vspace{1em}
I've attached my CV, which further details my skills and education. Please do not hesitate to reach out if you have any questions, and I look forward to the opportunity to speak to you further. I also have references available upon request. Thank you for your time and consideration.

\vspace{2em}
\noindent Sincerely,
\newline Jack

\vspace{2em}
\color{fade} \spacing{1}

\begin{tabbing}
Jack Symonds
\\ 6 Spectacle Ridge Rd. South Kent, CT 06785
\\ jacksymonds.js@gmail.com
\\ U.S. mobile: +1 \= (917)-573-3323
\\ U.K. mobile: +44 \= 7985 489436
\end{tabbing}
\end{document}
